A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-11 paper.Rmd
expanding

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with sex, social class, and legitimacy. Studies are performed 
separatelyfor neonatal  and post-neonatal mortality, and for winter and summer seasons, 
and in a joint analysis, the classification into endogenous and exogenous factors will be discussed.

@@

A variation of the Bourgeouis-Pichat (B-P) biometric analysis [@bp51a;
@bp51b] of infant mortality is suggested. B-P suggests that mortality
during the last eleven months of infancy follows a *uniform* distribution
given a specific transformation of time (age), the 
*log-cube ("BP") transformation*. We show the BP transformation is closely approximated by
an *exponential distribution*, 
that is, the *cumulative hazards function* is linear.
 The advantage is that estimation of exogeneous and
endogeneous infant mortality is easily performed with standard survival
analysis programs.

Central in the *Bourgeouis-Pichat (B-P)* model is the *log-cube* time transform $g$:

@@

```{r figgraph, fig.cap = "The log-cube transform g."}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 365, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, 
     col = "blue")
xat <- c(0, 28, 183, 365)
yat <- g(xat)
axis(1, at = xat)
axis(2, at = yat, labels = round(yat, 1), las = 1)
box()
lines(c(28, 28), c(0, g(28)), lty = 2, col = "darkgreen")
lines(c(0, 28), c(g(28), g(28)), lty = 2, col = "darkgreen")
##
lines(c(183, 183), c(0, g(183)), lty = 2, col = "darkgreen")
lines(c(0, 183), c(g(183), g(183)), lty = 2, col = "darkgreen")
##
lines(c(365, 365), c(0, g(365)), lty = 2, col = "darkgreen")
lines(c(0, 365), c(g(365), g(365)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
##abline(a = 0, b = 1, col = "red")
```

The B-P model specifies that, on the transformed age scale, the *exogeneous*
mortality distribution, conditional on death before age 365 days,  is *uniform*, that is, the 
survivor function of $T$ is *linear*.

\begin{equation}
S(t) = 1 - \alpha t, \, 0 < t < g(1).
\end{equation}

@@

It seems to fit infant mortality data very well, but one drawback with it is that the 
uniform distribution is not easy to work with in a survival analysis context.

However, we show that the BP transformation is well approximated by a model where the 
*cumulative hazards function* is linear instead, that is, an *exponential distribution*,
easy to use in standard survival regression analysis, often allowing for massive data reduction
through the statistical *sufficency principle*.

## Infant Mortality

@@

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
reason, whichever comes first. The first few lines (infants) are shown in Table 
\@ref(tab:infdat3).

If `event == FALSE`, then the infant died at age (in days) given by `exit`, otherwise
the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. For the
time being, no covariates are involved. So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
while infant No. 5 died at age 162 days.

@@

The BP procedure consists of two steps, to begin with:

1.   Apply the log-cube transformation to the age data, and

2.   plot the cumulative sum of the number of deaths vs. age.

The result is shown in Figure \@ref(fig:ggplot2), with the original, untransformed
data together with the transform, stretched to 365.

```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950."}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     col = "red", lwd = 2, xlab = "Age (days)",
     main = "After g-transform", ylab = "Cum. deaths")
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red")
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

The plot, disregarding the transform, is just a graph of the empirical *cumulative distribution function (CDF)*,
that is, one minus the *Kaplan-Meier estimate* of the *survivor function*.

### Post-neonatal mortality

We begin by studying mortality after age 28 days on the $g$-scale, *B-P* predicts that it is approximately *uniform*. 
The uniform distribution, however, is not very useful in a survival analysis context, so we will
investigate a close alternative, the *exponential * distribution. Modifying the *B-P* plot by using successive 
divisions by *risk set size* instead of sample size should result in an approximately linear graph if the 
exponential assumption is valid.

```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed post-neonatal mortality data."}
ginfdat <- age.window(infdat, c(28, 365))
ginfdat$enter <- g(ginfdat$enter)
ginfdat$exit <- g(ginfdat$exit)
grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
ser <- plot(fit, fn = "sur", fig = FALSE)
##plot(fit, col = "blue", axes = FALSE)
cum <- fit$hazards[[1]]
x <- cum[, 1]
ch <- cumsum(cum[, 2])
plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
axis(2, las = 1)
abline(h = 0)
box()
x <- ser[[1]][[1]][, 1]
y <- ser[[1]][[1]][, 2]
lines(x, 1 - y, col = "red", lty = 2)
legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeouis-Pichat)"), col = c("blue", "red"),
       lty = c(1, 2))
```

As is evident from Figure \@ref(fig:pneo), the difference between the the two curves is
negligible, and both are well approximated by a straight line, so we will from now on work with the 
assumption that the transformed post-neonatal data are *exponentially distributed*.
There is also a mathematical argument motivating the similarity: The cdf for an exponentially 
distributed random variable $T$ is

@@

So, the *hazard function* (which is the slope of the lines in Figure \@ref(fig:pneo))
is simply estimated by the ratio of the number of deaths to the total time of exposure:

@@

Is the sum of two components, *external* and *internal* causes. According to the *B-P*
theory, the external component is equal to the post-neonatal mortality, which contains 
only an external component.

@@

and then fit a piecewise constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
Figure \@ref(fig:pchneo).

```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 4}
neocuts <- g(c(0, 2, 4, 7, 11, 16, 22, 28))
fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", axes = FALSE)
axis(1)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

Let us take a closer look at what happens at the end of the neonatal interval, see 
Figure \@ref(fig:pchneo2).

```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 4}
##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
par(las = 1)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(5, 39))
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

A smooth transition to the post-neonatal mortality.

@@

and with $h_Y$ shown in Figure \@ref(fig:together), we get $h_T$ in Figure 
\@ref(fig:allnorm).

```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale."}
y <- hazards(fitall, cum = FALSE)$x
hy <- hazards(fitall, cum = FALSE)$y
te <- ginv(y)
hte <- dg(te) * hy
plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
     ylim = c(0, 0.002))
axis(1, at = ginv(pcuts))
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

@@

## Weather data

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
infdat <- readRDS("data/infdat3_year3.rds")
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

In accordance with the ideas of Bourgeouis-Pichat, we group the analyses into two parts, 
*neonatal* (first 28 days) och *postneonatal* mortality (days 29--365).

## Infant data

In the previous chapter we introduced data on infant mortality from the Umeå-Skellefteå
region 1895--1950, and here we introduce covariates of interest in the forthcoming  analyses, 
see Table

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

@@

## Postneonatal mortality

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.04, 0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

The seasonal pattern is once again similar to the one we found for infant mortality.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with sex, social class, and legitimacy. Studies are performed 
separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
and in a joint analysis, the classification into endogenous and exogenous factors will be discussed.

One important reason for studying neonatal and postneonatal mortality separately is the 
findings by Bourgeouis-Pichat [@bp51a;  @bp51b] about endogeneous and exogeneous mortality and the log-cube transform,
so we start by recollecting his results and formulate them in a modern statistical setting.

@@

Bourgeouis-Picat (*B-P*) suggests that mortality
during the last eleven months (28 days) of infancy follows a *uniform* distribution
given a specific transformation of time (age), the 
*log-cube transformation* $g$.

@@

```{r figgraph, fig.cap = "The log-cube transform g.", fig.height = 4}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 365, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, 
     col = "blue")
xat <- c(0, 28, 183, 365)
yat <- g(xat)
axis(1, at = xat)
axis(2, at = yat, labels = round(yat, 1), las = 1)
box()
lines(c(28, 28), c(0, g(28)), lty = 2, col = "darkgreen")
lines(c(0, 28), c(g(28), g(28)), lty = 2, col = "darkgreen")
##
lines(c(183, 183), c(0, g(183)), lty = 2, col = "darkgreen")
lines(c(0, 183), c(g(183), g(183)), lty = 2, col = "darkgreen")
##
lines(c(365, 365), c(0, g(365)), lty = 2, col = "darkgreen")
lines(c(0, 365), c(g(365), g(365)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
##abline(a = 0, b = 1, col = "red")
```

The *B-P* model specifies that, on the transformed age scale, the *exogeneous*
mortality distribution, conditional on death before age 365 days,  is *uniform*, that is, the 
survivor function of $T$ is *linear*,

\begin{equation}
S(t) = 1 - \alpha t, \, 0 \le t \le g(365).
\end{equation}

@@

Next we apply the model to our infant mortality data from the Umeå-Skellefteå region.

## Infant mortality, a first look

@@

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
reason, whichever comes first. The first few lines (infants) are shown in Table 
\@ref(tab:infdat3). Note that covariates are left out for the time being.

If `event == TRUE`, then the infant died at age (in days) given by `exit`, otherwise
the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. 
So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
while infant No. 5 died at age 162 days. We assume here that all infants are followed from birth,
that is, `enter == 0` in all cases (no left truncation).

@@

The BP procedure consists of three steps:

1.   Order the data set ascending by `exit`,

2.   apply the log-cube transformation to the age data, and

3.   plot the cumulative sum of the number of deaths vs. age.

The result is shown in Figure \@ref(fig:ggplot2), together with the original
data.  The transform data is stretced  stretched to 365 for easier comparison.

```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950."}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, 
     col = "red", lwd = 2, xlab = "Age (days)", lty = 2)
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red", lty = 2)
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

The curves are the empirical *cumulative distribution function (CDF)* expressed 
on the two scales, that is, one minus the *Kaplan-Meier [@km58] estimates* of the 
*survivor functions*.

### Postneonatal mortality

We begin by studying mortality after age 28 days on the $g$-scale, *B-P* predicts that it is approximately *uniform*. 
The uniform distribution, however, is not very useful in a survival analysis context, so we will
investigate a close alternative, the *exponential * distribution. Modifying the *B-P* plot by using successive 
divisions by current *risk set size* instead of original sample size should result in an approximately linear graph if the 
exponential assumption is valid.

```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed postneonatal mortality data."}
ginfdat <- age.window(infdat, c(28, 365))
ginfdat$enter <- g(ginfdat$enter)
ginfdat$exit <- g(ginfdat$exit)
grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
ser <- plot(fit, fn = "sur", fig = FALSE)
##plot(fit, col = "blue", axes = FALSE)
cum <- fit$hazards[[1]]
x <- cum[, 1]
ch <- cumsum(cum[, 2])
plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
axis(2, las = 1)
abline(h = 0)
box()
x <- ser[[1]][[1]][, 1]
y <- ser[[1]][[1]][, 2]
lines(x, 1 - y, col = "red", lty = 2)
legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeouis-Pichat)"), col = c("blue", "red"),
       lty = c(1, 2))
```

As is evident from Figure \@ref(fig:pneo), the difference between the the two curves is
negligible, and both are well approximated by a straight line, so we will from now on work with the 
assumption that the transformed postneonatal data are *exponentially distributed*.
There is also a mathematical argument motivating the similarity: The cdf for an exponentially 
distributed random variable $T$ is

@@

So, the *hazard function* (which is the slope of the lines in Figure \@ref(fig:pneo))
is simply estimated by the ratio of the *number of deaths* to the *total time of exposure*
(an *occurrence/exposure* rate):

@@

Is the sum of two components, *external* and *internal* causes. According to the *B-P*
theory, the external component is equal to the postneonatal mortality, which contains 
only an external component.

@@

and then fit a piecewise constant hazards model with cutpoints at $g(0, 2, 4, 7, 11, 16, 22, 28)$, see 
Figure \@ref(fig:pchneo).

```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 3}
neocuts <- g(c(0, 2, 4, 7, 11, 16, 22, 28))
fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", axes = FALSE)
axis(1)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

Let us take a closer look at what happens at the end of the neonatal interval, see 
Figure \@ref(fig:pchneo2). It shows a smooth transition towards the postneonatal 
(external) mortality.

```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 3}
##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
par(las = 1)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(5, 39))
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

@@

and with $h_Y$ shown in Figure \@ref(fig:together), we get $h_T$ in Figure 
\@ref(fig:allnorm) **(must be checked!)**.

```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale."}
y <- hazards(fitall, cum = FALSE, n.points = 3000)$x
hy <- hazards(fitall, cum = FALSE, n.points = 3000)$y
te <- ginv(y)
hte <- dg(te) * hy
plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
     ylim = c(0, 0.002)) 
     ##xlim = c(0, 28))
axis(1, at = ginv(pcuts))
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

@@

## Infant mortality, summary graphs

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
infdat <- readRDS("data/infdat3_year3.rds")
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
abline(h = 0)
box()
```

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "NMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
abline(h = 0)
box()
```

@@

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "PNMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

The seasonal pattern is once again similar to the one we found for infant mortality.
Next, the decline over the years in Figures \@ref(fig:longtrend) and \@ref(fig:ntrend).

```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
y <- trend$event / trend$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "IMR")
abline(h = 0)
```

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "NMR")
abline(h = 0)
```
